---
title: "DATA3888-BIOMED9"
format: html
editor: visual
---

## **Disease: Systemic lupus erythematosus (SLE)**

A chronic autoimmune disease where the immune system attacks healthy tissues, causing inflammation and damage to joints, skin, kidneys, heart, lungs, and the brain. Often triggered by a mix of genetics, environmental factors (e.g., UV light, viruses), and hormones.

## Datasets

-   **GSE88884**: This study characterized baseline gene expression and pharmacodynamic (PD)-induced changes in whole blood gene expression from two phase III, 52-week (W), randomized, placebo-controlled, double-blind studies of 1,760 SLE patients treated with the B cell activating factor (BAFF)-blocking IgG4 monoclonal antibody, tabalumab. Methods: Patient samples were obtained from ILLUMINATE-1 and -2 while control samples were from healthy donors.

    <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE88884>

-   **GSE72509**: RNA-seq of systemic lupus erythematosus (SLE) whole blood and healthy controls to determine the gene expression changes in these patients. <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE72509>

## Research questions

1.  Can whole-blood RNA-seq distinguish SLE patients from healthy controls?

## Innovation

1.  predict with clinical features: combine gene expression with age, sex, ...
    -   Question can be: Does combining gene expression and clinical features improve prediction compared with gene expression alone?

    -   make the calculator closer to real clinical use
2.  start with high-dimensional genes; reduce to a 5-20 gene panel; fit LASSO logistic regression or elastic net; build a compact calculator from the selected genes

## EDA

```{r}
library(GEOquery) 
library(R.utils)
library(reshape2)
library(ggplot2)
library(limma)
library(dplyr)
```

**GSE72509**

it uses RPKM instead of raw counts

-   **117 samples total**: **99 SLE** samples and **18 healthy controls**.

```{r}
gse <- getGEO("GSE72509")
gse <- gse[[1]]
pheno_data <- pData(gse)

getGEOSuppFiles("GSE72509", makeDirectory = TRUE, baseDir = ".")

counts_df <- read.delim(
  gzfile("GSE72509_SLE_RPKMs.txt.gz"),
  check.names = FALSE
)

dim(counts_df)
head(counts_df)

feature_cols <- c("ALIASES", "CHR", "START", "END", "STR", "COORDS",
                  "SIZE", "TYPE", "SYMBOL", "DESC", "FEATURE_ID")
feature_data <- counts_df[, feature_cols]

gene_id <- counts_df[[1]]
exprMat <- as.matrix(counts_df[, -c(1:11)])
rownames(exprMat) <- gene_id
mode(exprMat) <- "numeric"

dim(exprMat)
exprMat[1:5, 1:5]

gse$Outcome <- ifelse(grepl("SLE", gse$title), "SLE", "Control")
table(gse$Outcome)

library(preprocessCore)
log_exprMat <- log2(exprMat + 1)
summary(log_exprMat[, 1])

ggplot(melt(log_exprMat), aes(x=Var2, y=value)) +  
  geom_boxplot(outlier.colour="black", outlier.shape=16,outlier.size=0.5, notch=FALSE) +
  theme(axis.text.x = element_text(angle = 180, hjust = 1)) +
  labs (x = "Patient", y = "Expression value") +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90))

```

```{r}
cors <- cor(log_exprMat)
heatmap(cors)
```

```{r}
pca <- prcomp(t(log_exprMat))

df_toplot <- data.frame(gse$Outcome, 
                        PC1 = pca$x[,1], PC2 = pca$x[,2]  )

ggplot(df_toplot, aes(x = PC1, y = PC2, color = gse.Outcome)) + 
  geom_point(size = 4) + 
  theme_minimal()
```

```{r}
design <- model.matrix(~gse$Outcome, data = pheno_data)
fit <- lmFit(log_exprMat, design)
fit <- eBayes(fit)

library(DT)
tT <- topTable(fit, genelist=feature_data[,"SYMBOL"], n = Inf)
tT[1:100, ]

summary(tT$logFC)
```
