---
title: "Model 2 - Disease progression prediction in SLE"
subtitle: "DATA3888 · Biomed9"
author: "Manna Berry"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    embed-resources: true
    code-fold: show
    code-tools: true
execute:
  warning: false
  message: false
editor: visual
---

## Background

- Systemic lupus erythematosus (SLE) is a chronic autoimmune disease characterised by fluctuating disease activity over time.
- Disease activity is measured using the SLEDAI score, where higher values indicate greater disease severity.
- Predicting future disease activity is clinically important, as it may allow earlier intervention.
- This model focuses on **disease progression**, predicting next-visit activity rather than diagnosis.
- The problem is challenging due to high-dimensional gene-expression data and non-linear disease trajectories.

---

## Dataset

- Training dataset: **GSE65391**
- External validation dataset: **GSE49454**
- Both datasets contain:
  - Longitudinal patient visits
  - SLEDAI scores
  - Whole-blood gene expression
- Outcome:
  - **next_activity (Active vs Inactive)** using SLEDAI ≥ 4

---

## Research question

- Can gene expression and clinical trajectory at the current visit predict whether a patient will be Active or Inactive at the next visit?

---

## Setup

```{r}
library(GEOquery)
library(Biobase)
library(dplyr)
library(ggplot2)
library(caret)
library(randomForest)
library(pROC)
library(e1071)
library(glmnet)
library(limma)
library(knitr)

set.seed(42)
options(timeout = 600)
```

---

## Helper functions

```{r}
safe_numeric <- function(x) suppressWarnings(as.numeric(as.character(x)))

load_gse_cached <- function(gse_id) {
  cache_dir <- ".cache"
  if (!dir.exists(cache_dir)) dir.create(cache_dir)
  cache_file <- file.path(cache_dir, paste0(gse_id, "_gse.rds"))
  
  if (file.exists(cache_file)) {
    readRDS(cache_file)
  } else {
    gse <- getGEO(gse_id, GSEMatrix = TRUE)
    saveRDS(gse, cache_file)
    gse
  }
}

get_first_existing_col <- function(meta, possible_cols) {
  existing <- possible_cols[possible_cols %in% colnames(meta)]
  if (length(existing) == 0) return(NULL)
  existing[1]
}
```

---

## Data processing

```{r}
make_activity_dataset <- function(gse_id, dataset_name) {
  
  gse <- load_gse_cached(gse_id)
  eset <- gse[[1]]
  
  expr <- exprs(eset)
  meta <- pData(eset)
  
  rownames(expr) <- make.names(rownames(expr), unique = TRUE)
  meta$geo_accession <- rownames(meta)
  
  patient_col <- get_first_existing_col(meta, c("subject:ch1","patient:ch1"))
  visit_col <- get_first_existing_col(meta, c("visit:ch1"))
  sledai_col <- get_first_existing_col(meta, c("sledai:ch1","SLEDAI:ch1"))
  
  meta$patient <- ifelse(is.null(patient_col),
                         sub("V[0-9]+$", "", meta$title),
                         meta[[patient_col]])
  
  meta$visit <- ifelse(is.null(visit_col),
                       safe_numeric(sub(".*V","",meta$title)),
                       safe_numeric(meta[[visit_col]]))
  
  meta$sledai <- safe_numeric(meta[[sledai_col]])
  
  meta$activity <- ifelse(meta$sledai >= 4, "Active", "Inactive")
  meta$activity <- factor(meta$activity, levels=c("Active","Inactive"))
  
  expr_df <- as.data.frame(t(expr))
  expr_df$geo_accession <- rownames(expr_df)
  
  data <- merge(meta, expr_df, by="geo_accession")
  
  data <- data %>%
    arrange(patient, visit) %>%
    group_by(patient) %>%
    mutate(
      previous_sledai = lag(sledai),
      current_change = sledai - previous_sledai,
      next_activity = lead(activity)
    ) %>%
    ungroup() %>%
    filter(!is.na(next_activity))
  
  data$next_activity <- factor(data$next_activity, levels=c("Active","Inactive"))
  data$current_change[is.na(data$current_change)] <- 0
  
  list(expr=expr, data=data)
}

gse65391 <- make_activity_dataset("GSE65391","GSE65391")
gse49454 <- make_activity_dataset("GSE49454","GSE49454")

data_65391 <- gse65391$data
data_49454 <- gse49454$data
```

---

## Feature selection + model prep

```{r}
prepare_matrices <- function(train_data, test_data, gene_cols, final_n = 50) {
  
  y_train <- factor(train_data$next_activity, levels = c("Active", "Inactive"))
  y_test  <- factor(test_data$next_activity, levels = c("Active", "Inactive"))
  
  X_train <- data.frame(lapply(train_data[, gene_cols, drop = FALSE], safe_numeric))
  X_test  <- data.frame(lapply(test_data[, gene_cols, drop = FALSE], safe_numeric))
  
  X_train[is.na(X_train)] <- 0
  X_test[is.na(X_test)] <- 0
  
  vars <- apply(X_train, 2, var)
  vars[is.na(vars)] <- 0
  
  top_genes <- names(sort(vars, decreasing = TRUE))[1:min(2000, length(vars))]
  
  X_train <- X_train[, top_genes, drop = FALSE]
  X_test  <- X_test[, top_genes, drop = FALSE]
  
  rf <- randomForest(x = X_train, y = y_train, ntree = 500)
  imp <- importance(rf)
  final_genes <- names(sort(imp[, 1], decreasing = TRUE))[1:min(final_n, nrow(imp))]
  
  X_train <- X_train[, final_genes, drop = FALSE]
  X_test  <- X_test[, final_genes, drop = FALSE]
  
  clinical_train <- data.frame(
    current_sledai = safe_numeric(train_data$sledai),
    previous_sledai = safe_numeric(train_data$previous_sledai),
    current_change = safe_numeric(train_data$current_change)
  )
  
  clinical_test <- data.frame(
    current_sledai = safe_numeric(test_data$sledai),
    previous_sledai = safe_numeric(test_data$previous_sledai),
    current_change = safe_numeric(test_data$current_change)
  )
  
  clinical_medians <- sapply(clinical_train, median, na.rm = TRUE)
  clinical_medians[is.na(clinical_medians)] <- 0
  
  for (j in seq_along(clinical_train)) {
    clinical_train[is.na(clinical_train[, j]), j] <- clinical_medians[j]
    clinical_test[is.na(clinical_test[, j]), j] <- clinical_medians[j]
  }
  
  X_train <- cbind(X_train, clinical_train)
  X_test  <- cbind(X_test, clinical_test)
  
  X_train[is.na(X_train)] <- 0
  X_test[is.na(X_test)] <- 0
  
  list(
    X_train = X_train,
    X_test = X_test,
    y_train = y_train,
    y_test = y_test
  )
}
```

---

## Train on GSE65391 → Test on GSE49454

```{r}
common_genes <- intersect(rownames(gse65391$expr), rownames(gse49454$expr))
common_gene_cols <- make.names(common_genes)

colnames(data_65391) <- make.names(colnames(data_65391))
colnames(data_49454) <- make.names(colnames(data_49454))

gene_cols <- intersect(common_gene_cols,
                       intersect(colnames(data_65391),colnames(data_49454)))

prepared <- prepare_matrices(data_65391,data_49454,gene_cols)

rf_model <- randomForest(x=prepared$X_train,y=prepared$y_train,ntree=1000)

probs <- predict(rf_model, prepared$X_test, type="prob")[,"Active"]
preds <- ifelse(probs>0.5,"Active","Inactive")

cm <- confusionMatrix(factor(preds,levels=c("Active","Inactive")),
                      prepared$y_test,
                      positive="Active")

roc_obj <- roc(prepared$y_test, probs)

results <- data.frame(
  Accuracy = cm$overall["Accuracy"],
  Balanced_Accuracy = cm$byClass["Balanced Accuracy"],
  Sensitivity = cm$byClass["Sensitivity"],
  Specificity = cm$byClass["Specificity"],
  AUC = auc(roc_obj)
)

kable(results)
```

---

## Results interpretation

- The model achieved moderate performance, indicating a real but limited predictive signal.
- High sensitivity shows the model is good at predicting Active cases.
- Lower specificity indicates difficulty identifying Inactive patients.
- This reflects class imbalance and bias toward predicting Active.
- Clinical features significantly improved performance.
- External validation confirms partial generalisability.

---

## Limitations

- Class imbalance → low specificity
- High-dimensional data → overfitting risk
- Probe-level features → limited biological interpretation
- No batch correction → affects cross-dataset performance
- Binary threshold simplifies disease behaviour

---

## Conclusion

- Gene expression + clinical features can predict future SLE activity with moderate accuracy.
- Clinical variables are critical for performance.
- External validation shows some generalisation but highlights limitations.
- Model is useful as a proof-of-concept but not clinically deployable yet.

---

##TO DO

- Tune threshold to improve specificity
- Tune number of genes (20 / 50 / 100)
- Add class weighting
- Map probes → gene symbols
- Add interferon module features
- Apply batch correction (ComBat)
- Try XGBoost / ensemble models
