---
title: "Model 2 - Disease progression prediction in SLE"
subtitle: "DATA3888 · Biomed9"
author: "Udit Samant"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    embed-resources: true
    code-fold: show
    code-tools: true
    df-print: paged
    fig-width: 8
    fig-height: 5
    fig-align: center
    highlight-style: github
    smooth-scroll: true
execute:
  warning: false
  message: false
editor: visual
---

## Background

### Systemic lupus erythematosus

\[WRITE HERE\]

### Disease progression over time

\[WRITE HERE\]

## Dataset

This analysis uses two longitudinal SLE gene-expression datasets: **GSE65391** and **GSE49454**. Both contain repeated patient visits, SLEDAI disease activity scores, and whole-blood gene-expression measurements. The target variable is the patient's **next-visit disease activity**, classified as Active or Inactive using SLEDAI.

## Research question

Can current-visit gene expression and recent clinical trajectory predict whether SLE disease activity will be Active or Inactive at the next recorded visit?

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(GEOquery)
library(Biobase)
library(dplyr)
library(ggplot2)
library(caret)
library(randomForest)
library(pROC)
library(e1071)
library(glmnet)
library(limma)
library(knitr)

set.seed(42)
options(timeout = 600)
```

## Helper functions

```{r}
#| label: helper-functions

safe_numeric <- function(x) {
  suppressWarnings(as.numeric(as.character(x)))
}

get_first_existing_col <- function(meta, possible_cols) {
  existing <- possible_cols[possible_cols %in% colnames(meta)]
  if (length(existing) == 0) return(NULL)
  existing[1]
}

load_gse_cached <- function(gse_id) {
  cache_dir <- ".cache"
  if (!dir.exists(cache_dir)) dir.create(cache_dir)
  
  cache_file <- file.path(cache_dir, paste0(gse_id, "_gse.rds"))
  
  if (file.exists(cache_file)) {
    gse <- readRDS(cache_file)
  } else {
    gse <- getGEO(gse_id, GSEMatrix = TRUE)
    saveRDS(gse, cache_file)
  }
  
  gse
}

find_symbol_column <- function(feature_data) {
  possible_cols <- c(
    "Gene Symbol", "gene_assignment", "Symbol", "SYMBOL",
    "Gene.symbol", "GENE_SYMBOL", "ILMN_Gene"
  )
  
  existing <- possible_cols[possible_cols %in% colnames(feature_data)]
  
  if (length(existing) == 0) return(NULL)
  existing[1]
}

clean_gene_symbol <- function(x) {
  x <- as.character(x)
  x <- sub("///.*", "", x)
  x <- sub("//.*", "", x)
  x <- sub(";.*", "", x)
  x <- sub(",.*", "", x)
  x <- trimws(x)
  x[x == "" | is.na(x)] <- NA
  x
}

collapse_expression_to_symbols <- function(expr, feature_data) {
  
  symbol_col <- find_symbol_column(feature_data)
  
  if (is.null(symbol_col)) {
    message("No gene-symbol column found. Using probe IDs as features.")
    expr_symbol <- expr
    rownames(expr_symbol) <- make.names(rownames(expr_symbol), unique = TRUE)
    return(expr_symbol)
  }
  
  symbols <- clean_gene_symbol(feature_data[[symbol_col]])
  
  keep <- !is.na(symbols)
  expr <- expr[keep, , drop = FALSE]
  symbols <- symbols[keep]
  
  symbols <- make.names(symbols)
  
  expr_df <- as.data.frame(expr)
  expr_df$gene_symbol <- symbols
  
  collapsed <- expr_df %>%
    group_by(gene_symbol) %>%
    summarise(across(where(is.numeric), mean, na.rm = TRUE), .groups = "drop")
  
  collapsed_mat <- as.matrix(collapsed[, -1])
  rownames(collapsed_mat) <- collapsed$gene_symbol
  
  collapsed_mat
}

make_activity_dataset <- function(gse_id, dataset_name) {
  
  gse <- load_gse_cached(gse_id)
  
  eset <- gse[[1]]
  expr_raw <- exprs(eset)
  meta <- pData(eset)
  fdat <- fData(eset)
  
  expr <- collapse_expression_to_symbols(expr_raw, fdat)
  
  meta$geo_accession <- rownames(meta)
  
  patient_col <- get_first_existing_col(
    meta,
    c("subject:ch1", "patient:ch1", "patient_id:ch1", "id:ch1")
  )
  
  visit_col <- get_first_existing_col(
    meta,
    c("visit:ch1", "time:ch1", "week:ch1")
  )
  
  sledai_col <- get_first_existing_col(
    meta,
    c("sledai:ch1", "SLEDAI:ch1", "sledai", "SLEDAI")
  )
  
  if (is.null(sledai_col)) {
    stop(paste("Could not find SLEDAI column for", gse_id))
  }
  
  if (is.null(patient_col)) {
    meta$patient <- sub("V[0-9]+$", "", meta$title)
  } else {
    meta$patient <- meta[[patient_col]]
  }
  
  if (is.null(visit_col)) {
    meta$visit <- safe_numeric(sub(".*V", "", meta$title))
  } else {
    meta$visit <- safe_numeric(meta[[visit_col]])
  }
  
  meta$sledai <- safe_numeric(meta[[sledai_col]])
  meta$activity <- ifelse(meta$sledai >= 4, "Active", "Inactive")
  meta$activity <- factor(meta$activity, levels = c("Active", "Inactive"))
  meta$dataset <- dataset_name
  
  expr_df <- as.data.frame(t(expr))
  expr_df$geo_accession <- rownames(expr_df)
  
  data <- merge(meta, expr_df, by = "geo_accession")
  
  data <- data %>%
    filter(!is.na(patient), !is.na(visit), !is.na(sledai), !is.na(activity)) %>%
    arrange(patient, visit) %>%
    group_by(patient) %>%
    mutate(
      previous_sledai = lag(sledai),
      current_change = sledai - previous_sledai,
      next_sledai = lead(sledai),
      next_activity = lead(activity)
    ) %>%
    ungroup() %>%
    filter(!is.na(next_activity))
  
  data$next_activity <- factor(data$next_activity, levels = c("Active", "Inactive"))
  data$current_change[is.na(data$current_change)] <- 0
  data$previous_sledai[is.na(data$previous_sledai)] <- data$sledai[is.na(data$previous_sledai)]
  
  list(
    expr = expr,
    meta = meta,
    data = data,
    expr_df = expr_df
  )
}

patient_split <- function(data, patient_col = "patient", outcome_col = "next_activity",
                          p = 0.7, seed = 42) {
  
  set.seed(seed)
  
  patient_summary <- data %>%
    group_by(.data[[patient_col]]) %>%
    summarise(
      outcome_majority = names(sort(table(.data[[outcome_col]]), decreasing = TRUE))[1],
      .groups = "drop"
    )
  
  train_patients_idx <- createDataPartition(
    factor(patient_summary$outcome_majority),
    p = p,
    list = FALSE
  )
  
  train_patients <- patient_summary[[patient_col]][train_patients_idx]
  
  train_idx <- which(data[[patient_col]] %in% train_patients)
  test_idx <- setdiff(seq_len(nrow(data)), train_idx)
  
  list(train_idx = train_idx, test_idx = test_idx)
}

find_best_threshold <- function(y_true, probs) {
  
  thresholds <- seq(0.1, 0.9, by = 0.01)
  
  scores <- sapply(thresholds, function(t) {
    
    pred <- factor(
      ifelse(probs >= t, "Active", "Inactive"),
      levels = c("Active", "Inactive")
    )
    
    cm <- confusionMatrix(pred, y_true, positive = "Active")
    as.numeric(cm$byClass["Balanced Accuracy"])
  })
  
  thresholds[which.max(scores)]
}

evaluate_binary_model <- function(pred_class, pred_prob, y_test, model_name) {
  
  pred_class <- factor(pred_class, levels = c("Active", "Inactive"))
  y_test <- factor(y_test, levels = c("Active", "Inactive"))
  
  cm <- confusionMatrix(pred_class, y_test, positive = "Active")
  
  roc_obj <- roc(
    response = y_test,
    predictor = pred_prob,
    levels = c("Inactive", "Active"),
    quiet = TRUE
  )
  
  data.frame(
    Model = model_name,
    Accuracy = as.numeric(cm$overall["Accuracy"]),
    Balanced_Accuracy = as.numeric(cm$byClass["Balanced Accuracy"]),
    Sensitivity = as.numeric(cm$byClass["Sensitivity"]),
    Specificity = as.numeric(cm$byClass["Specificity"]),
    AUC = as.numeric(auc(roc_obj))
  )
}

make_class_weights <- function(y) {
  tab <- table(y)
  weights <- sum(tab) / (length(tab) * tab)
  weights
}

make_observation_weights <- function(y) {
  cw <- make_class_weights(y)
  as.numeric(cw[as.character(y)])
}

best_tune_predictions <- function(model) {
  pred <- model$pred
  best <- model$bestTune
  
  for (nm in names(best)) {
    pred <- pred[pred[[nm]] == best[[nm]], ]
  }
  
  pred
}

add_module_scores <- function(X_train_full, X_test_full) {
  
  interferon_genes <- make.names(c(
    "IFI44", "IFI44L", "MX1", "ISG15", "OAS1", "OAS2",
    "OAS3", "IFIT1", "IFIT2", "IFIT3", "RSAD2", "IRF7",
    "STAT1", "XAF1", "USP18", "EPSTI1"
  ))
  
  inflammation_genes <- make.names(c(
    "IL1B", "TNF", "CXCL10", "CXCL11", "CCL2", "CCR1",
    "NFKB1", "NFKBIA", "S100A8", "S100A9"
  ))
  
  bcell_genes <- make.names(c(
    "CD79A", "CD79B", "MS4A1", "BANK1", "BLK", "TNFRSF13C"
  ))
  
  module_score <- function(X, genes) {
    present <- intersect(genes, colnames(X))
    if (length(present) == 0) {
      return(rep(0, nrow(X)))
    }
    rowMeans(X[, present, drop = FALSE], na.rm = TRUE)
  }
  
  train_scores <- data.frame(
    IFN_score = module_score(X_train_full, interferon_genes),
    Inflammation_score = module_score(X_train_full, inflammation_genes),
    Bcell_score = module_score(X_train_full, bcell_genes)
  )
  
  test_scores <- data.frame(
    IFN_score = module_score(X_test_full, interferon_genes),
    Inflammation_score = module_score(X_test_full, inflammation_genes),
    Bcell_score = module_score(X_test_full, bcell_genes)
  )
  
  list(
    train = train_scores,
    test = test_scores,
    ifn_genes_found = intersect(interferon_genes, colnames(X_train_full)),
    inflammation_genes_found = intersect(inflammation_genes, colnames(X_train_full)),
    bcell_genes_found = intersect(bcell_genes, colnames(X_train_full))
  )
}

prepare_matrices <- function(train_data, test_data, gene_cols,
                             outcome_col = "next_activity",
                             include_clinical = TRUE,
                             include_modules = TRUE,
                             top_var_n = 2000,
                             top_limma_n = 300,
                             final_n = 50) {
  
  y_train <- factor(train_data[[outcome_col]], levels = c("Active", "Inactive"))
  y_test  <- factor(test_data[[outcome_col]], levels = c("Active", "Inactive"))
  
  X_train_full <- train_data[, gene_cols, drop = FALSE]
  X_test_full  <- test_data[, gene_cols, drop = FALSE]
  
  X_train_full <- data.frame(lapply(X_train_full, safe_numeric))
  X_test_full  <- data.frame(lapply(X_test_full, safe_numeric))
  
  keep <- sapply(
    X_train_full,
    function(x) {
      gene_var <- var(x, na.rm = TRUE)
      !all(is.na(x)) && is.finite(gene_var) && gene_var > 0
    }
  )
  
  X_train_full <- X_train_full[, keep, drop = FALSE]
  X_test_full  <- X_test_full[, keep, drop = FALSE]
  
  train_medians <- sapply(X_train_full, median, na.rm = TRUE)
  
  for (j in seq_along(X_train_full)) {
    X_train_full[is.na(X_train_full[, j]), j] <- train_medians[j]
    X_test_full[is.na(X_test_full[, j]), j] <- train_medians[j]
  }
  
  module_info <- NULL
  
  if (include_modules) {
    module_info <- add_module_scores(X_train_full, X_test_full)
  }
  
  gene_vars <- sapply(X_train_full, var, na.rm = TRUE)
  top_var_n <- min(top_var_n, length(gene_vars))
  var_genes <- names(sort(gene_vars, decreasing = TRUE))[1:top_var_n]
  
  X_train_var <- X_train_full[, var_genes, drop = FALSE]
  X_test_var  <- X_test_full[, var_genes, drop = FALSE]
  
  design <- model.matrix(~ y_train)
  fit <- eBayes(lmFit(t(as.matrix(X_train_var)), design))
  limma_tab <- topTable(fit, coef = 2, number = Inf, sort.by = "P")
  
  top_limma_n <- min(top_limma_n, nrow(limma_tab))
  limma_genes <- rownames(limma_tab)[1:top_limma_n]
  
  X_train_limma <- X_train_var[, limma_genes, drop = FALSE]
  
  class_weights_rf <- make_class_weights(y_train)
  
  rf_rank_fit <- randomForest(
    x = X_train_limma,
    y = y_train,
    ntree = 500,
    mtry = max(1, floor(sqrt(ncol(X_train_limma)))),
    classwt = class_weights_rf,
    importance = TRUE
  )
  
  rf_imp <- importance(rf_rank_fit, type = 2)
  rf_rank <- sort(rf_imp[, "MeanDecreaseGini"], decreasing = TRUE)
  
  final_n <- min(final_n, length(rf_rank))
  final_genes <- names(rf_rank)[1:final_n]
  
  X_train <- X_train_var[, final_genes, drop = FALSE]
  X_test  <- X_test_var[, final_genes, drop = FALSE]
  
  if (include_modules) {
    X_train <- cbind(X_train, module_info$train)
    X_test  <- cbind(X_test, module_info$test)
  }
  
  if (include_clinical) {
    
    clinical_train <- data.frame(
      current_sledai = safe_numeric(train_data$sledai),
      previous_sledai = safe_numeric(train_data$previous_sledai),
      current_change = safe_numeric(train_data$current_change),
      visit = safe_numeric(train_data$visit)
    )
    
    clinical_test <- data.frame(
      current_sledai = safe_numeric(test_data$sledai),
      previous_sledai = safe_numeric(test_data$previous_sledai),
      current_change = safe_numeric(test_data$current_change),
      visit = safe_numeric(test_data$visit)
    )
    
    clinical_medians <- sapply(clinical_train, median, na.rm = TRUE)
    
    for (j in seq_along(clinical_train)) {
      clinical_train[is.na(clinical_train[, j]), j] <- clinical_medians[j]
      clinical_test[is.na(clinical_test[, j]), j] <- clinical_medians[j]
    }
    
    X_train <- cbind(X_train, clinical_train)
    X_test  <- cbind(X_test, clinical_test)
  }
  
  preproc <- preProcess(X_train, method = c("center", "scale"))
  X_train_scaled <- predict(preproc, X_train)
  X_test_scaled  <- predict(preproc, X_test)
  
  colnames(X_train_scaled) <- make.names(colnames(X_train_scaled))
  colnames(X_test_scaled)  <- make.names(colnames(X_test_scaled))
  
  list(
    X_train = X_train_scaled,
    X_test = X_test_scaled,
    y_train = y_train,
    y_test = y_test,
    final_genes = final_genes,
    limma_table = limma_tab,
    rf_ranking = rf_rank,
    module_info = module_info
  )
}

train_and_evaluate_models <- function(prepared, label_prefix = "Model") {
  
  X_train <- prepared$X_train
  X_test  <- prepared$X_test
  y_train <- prepared$y_train
  y_test  <- prepared$y_test
  
  ctrl <- trainControl(
    method = "repeatedcv",
    number = 5,
    repeats = 2,
    classProbs = TRUE,
    summaryFunction = twoClassSummary,
    savePredictions = "final"
  )
  
  class_weights <- make_class_weights(y_train)
  obs_weights <- make_observation_weights(y_train)
  
  set.seed(42)
  rf_model <- randomForest(
    x = X_train,
    y = y_train,
    ntree = 1000,
    mtry = max(1, floor(sqrt(ncol(X_train)))),
    classwt = class_weights,
    importance = TRUE
  )
  
  rf_prob <- predict(rf_model, X_test, type = "prob")[, "Active"]
  rf_pred <- factor(
    ifelse(rf_prob >= 0.5, "Active", "Inactive"),
    levels = c("Active", "Inactive")
  )
  
  rf_perf <- evaluate_binary_model(
    rf_pred,
    rf_prob,
    y_test,
    paste(label_prefix, "Random Forest")
  )
  
  set.seed(42)
  svm_model <- train(
    x = X_train,
    y = y_train,
    method = "svmRadial",
    metric = "ROC",
    trControl = ctrl,
    tuneGrid = expand.grid(
      sigma = c(0.001, 0.005, 0.01),
      C = c(0.5, 1, 2, 4)
    ),
    class.weights = class_weights
  )
  
  svm_cv <- best_tune_predictions(svm_model)
  best_thresh_svm <- find_best_threshold(svm_cv$obs, svm_cv$Active)
  
  svm_prob <- predict(svm_model, X_test, type = "prob")[, "Active"]
  svm_pred <- factor(
    ifelse(svm_prob >= best_thresh_svm, "Active", "Inactive"),
    levels = c("Active", "Inactive")
  )
  
  svm_perf <- evaluate_binary_model(
    svm_pred,
    svm_prob,
    y_test,
    paste(label_prefix, "SVM")
  )
  
  set.seed(42)
  glmnet_model <- train(
    x = X_train,
    y = y_train,
    method = "glmnet",
    metric = "ROC",
    weights = obs_weights,
    trControl = ctrl,
    tuneGrid = expand.grid(
      alpha = c(0, 0.25, 0.5, 0.75, 1),
      lambda = seq(0.001, 0.05, length = 12)
    )
  )
  
  glmnet_cv <- best_tune_predictions(glmnet_model)
  best_thresh_glmnet <- find_best_threshold(glmnet_cv$obs, glmnet_cv$Active)
  
  glmnet_prob <- predict(glmnet_model, X_test, type = "prob")[, "Active"]
  glmnet_pred <- factor(
    ifelse(glmnet_prob >= best_thresh_glmnet, "Active", "Inactive"),
    levels = c("Active", "Inactive")
  )
  
  glmnet_perf <- evaluate_binary_model(
    glmnet_pred,
    glmnet_prob,
    y_test,
    paste(label_prefix, "Elastic Net")
  )
  
  performance <- rbind(rf_perf, svm_perf, glmnet_perf)
  performance[, -1] <- round(performance[, -1], 3)
  
  list(
    performance = performance,
    rf_model = rf_model,
    svm_model = svm_model,
    glmnet_model = glmnet_model,
    best_thresh_svm = best_thresh_svm,
    best_thresh_glmnet = best_thresh_glmnet
  )
}
```

## Loading the data

```{r}
#| label: load-datasets
#| message: false
#| warning: false

gse65391 <- make_activity_dataset("GSE65391", "GSE65391")
gse49454 <- make_activity_dataset("GSE49454", "GSE49454")

data_65391 <- gse65391$data
data_49454 <- gse49454$data

dim(data_65391)
dim(data_49454)

table(data_65391$next_activity)
table(data_49454$next_activity)
```

## Common gene alignment

```{r}
#| label: common-genes

genes_65391 <- rownames(gse65391$expr)
genes_49454 <- rownames(gse49454$expr)

common_genes <- intersect(genes_65391, genes_49454)

length(common_genes)

common_gene_cols <- make.names(common_genes)

colnames(data_65391) <- make.names(colnames(data_65391))
colnames(data_49454) <- make.names(colnames(data_49454))

gene_cols <- intersect(
  common_gene_cols,
  intersect(colnames(data_65391), colnames(data_49454))
)

length(gene_cols)
```

## Exploratory data analysis

### Next-visit activity class balance

```{r}
#| label: next-activity-class-balance

class_balance <- rbind(
  data.frame(
    Dataset = "GSE65391",
    as.data.frame(table(data_65391$next_activity))
  ),
  data.frame(
    Dataset = "GSE49454",
    as.data.frame(table(data_49454$next_activity))
  )
)

colnames(class_balance) <- c("Dataset", "Next_Activity", "Count")

class_balance

ggplot(class_balance, aes(x = Next_Activity, y = Count, fill = Dataset)) +
  geom_col(position = "dodge") +
  labs(
    title = "Next-visit disease activity class balance",
    x = "Next-visit activity",
    y = "Number of visits"
  ) +
  theme_minimal()
```

### SLEDAI distribution

```{r}
#| label: sledai-distribution

sledai_df <- rbind(
  data.frame(Dataset = "GSE65391", sledai = data_65391$sledai),
  data.frame(Dataset = "GSE49454", sledai = data_49454$sledai)
)

ggplot(sledai_df, aes(x = sledai, fill = Dataset)) +
  geom_histogram(binwidth = 2, alpha = 0.6, position = "identity") +
  labs(
    title = "Distribution of current SLEDAI scores",
    x = "SLEDAI score",
    y = "Count"
  ) +
  theme_minimal()
```

## Feature analysis and extraction

Feature selection was performed using training data only. The pipeline combines variance filtering, limma ranking, and random forest importance. In addition to individual genes, biologically informed module scores are added, including an interferon signature score.

## Internal validation using patient-level split

```{r}
#| label: internal-patient-split

split_65391 <- patient_split(data_65391, p = 0.7, seed = 42)

train_65391 <- data_65391[split_65391$train_idx, ]
test_65391  <- data_65391[split_65391$test_idx, ]

table(train_65391$next_activity)
table(test_65391$next_activity)
```

### Gene-only feature panel

```{r}
#| label: prepare-gene-only-internal

gene_only_internal <- prepare_matrices(
  train_data = train_65391,
  test_data = test_65391,
  gene_cols = gene_cols,
  include_clinical = FALSE,
  include_modules = FALSE,
  top_var_n = 2000,
  top_limma_n = 300,
  final_n = 50
)

dim(gene_only_internal$X_train)
dim(gene_only_internal$X_test)

head(gene_only_internal$final_genes, 20)
```

### Gene + module + clinical feature panel

```{r}
#| label: prepare-gene-module-clinical-internal

gene_module_clinical_internal <- prepare_matrices(
  train_data = train_65391,
  test_data = test_65391,
  gene_cols = gene_cols,
  include_clinical = TRUE,
  include_modules = TRUE,
  top_var_n = 2000,
  top_limma_n = 300,
  final_n = 50
)

dim(gene_module_clinical_internal$X_train)
dim(gene_module_clinical_internal$X_test)

gene_module_clinical_internal$module_info$ifn_genes_found
```

### Top selected genes

```{r}
#| label: top-selected-genes

top_rf <- head(gene_only_internal$rf_ranking, 20)

top_rf_df <- data.frame(
  gene = factor(names(top_rf), levels = rev(names(top_rf))),
  importance = as.numeric(top_rf)
)

ggplot(top_rf_df, aes(x = importance, y = gene)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Top 20 selected genes by random forest importance",
    x = "Mean decrease in Gini impurity",
    y = "Gene"
  ) +
  theme_minimal()
```

## Modelling

Class imbalance is handled using class weights rather than downsampling. This keeps all real observations while penalising misclassification of the minority class more strongly. SVM and elastic-net thresholds are optimised using cross-validation predictions to maximise balanced accuracy.

### Internal validation results

```{r}
#| label: internal-validation-results

internal_gene_only_results <- train_and_evaluate_models(
  gene_only_internal,
  label_prefix = "Internal gene-only"
)

internal_gene_module_clinical_results <- train_and_evaluate_models(
  gene_module_clinical_internal,
  label_prefix = "Internal gene + module + clinical"
)

internal_performance <- rbind(
  internal_gene_only_results$performance,
  internal_gene_module_clinical_results$performance
)

kable(
  internal_performance,
  caption = "Internal patient-level validation performance using GSE65391"
)

internal_gene_only_results$best_thresh_svm
internal_gene_only_results$best_thresh_glmnet
internal_gene_module_clinical_results$best_thresh_svm
internal_gene_module_clinical_results$best_thresh_glmnet
```

\[WRITE HERE: Compare gene-only versus gene + module + clinical results.\]

## Cross-dataset external validation

Models are trained on GSE65391 and externally evaluated on GSE49454 using common genes only. This tests whether the predictive signal generalises across an independent cohort.

```{r}
#| label: external-validation-prepare

external_gene_only <- prepare_matrices(
  train_data = data_65391,
  test_data = data_49454,
  gene_cols = gene_cols,
  include_clinical = FALSE,
  include_modules = FALSE,
  top_var_n = 2000,
  top_limma_n = 300,
  final_n = 50
)

external_gene_module_clinical <- prepare_matrices(
  train_data = data_65391,
  test_data = data_49454,
  gene_cols = gene_cols,
  include_clinical = TRUE,
  include_modules = TRUE,
  top_var_n = 2000,
  top_limma_n = 300,
  final_n = 50
)

table(external_gene_only$y_train)
table(external_gene_only$y_test)

external_gene_module_clinical$module_info$ifn_genes_found
```

```{r}
#| label: external-validation-results

external_gene_only_results <- train_and_evaluate_models(
  external_gene_only,
  label_prefix = "External gene-only"
)

external_gene_module_clinical_results <- train_and_evaluate_models(
  external_gene_module_clinical,
  label_prefix = "External gene + module + clinical"
)

external_performance <- rbind(
  external_gene_only_results$performance,
  external_gene_module_clinical_results$performance
)

kable(
  external_performance,
  caption = "External validation performance: trained on GSE65391 and tested on GSE49454"
)

external_gene_only_results$best_thresh_svm
external_gene_only_results$best_thresh_glmnet
external_gene_module_clinical_results$best_thresh_svm
external_gene_module_clinical_results$best_thresh_glmnet
```

\[WRITE HERE: External validation may be lower due to cohort, platform, and preprocessing differences.\]

## Combined dataset sensitivity analysis

As a secondary analysis, the two datasets are combined to increase sample size. This result should be interpreted cautiously because combining datasets can introduce batch effects.

```{r}
#| label: combined-dataset

shared_cols <- intersect(colnames(data_65391), colnames(data_49454))

combined_data <- rbind(
  data_65391[, shared_cols],
  data_49454[, shared_cols]
)

combined_data$dataset <- factor(combined_data$dataset)

table(combined_data$dataset)
table(combined_data$next_activity)
```

```{r}
#| label: combined-patient-split

split_combined <- patient_split(combined_data, p = 0.7, seed = 42)

train_combined <- combined_data[split_combined$train_idx, ]
test_combined  <- combined_data[split_combined$test_idx, ]

combined_gene_module_clinical <- prepare_matrices(
  train_data = train_combined,
  test_data = test_combined,
  gene_cols = gene_cols,
  include_clinical = TRUE,
  include_modules = TRUE,
  top_var_n = 2000,
  top_limma_n = 300,
  final_n = 50
)

combined_results <- train_and_evaluate_models(
  combined_gene_module_clinical,
  label_prefix = "Combined gene + module + clinical"
)

combined_performance <- combined_results$performance

kable(
  combined_performance,
  caption = "Combined dataset patient-level validation performance"
)

combined_results$best_thresh_svm
combined_results$best_thresh_glmnet
```

## Overall performance summary

```{r}
#| label: overall-performance-summary

overall_performance <- rbind(
  internal_performance,
  external_performance,
  combined_performance
)

kable(
  overall_performance,
  caption = "Overall comparison of internal, external, and combined modelling strategies"
)
```

## Results interpretation

\[WRITE HERE\]

This model returns to the next-visit Active versus Inactive target, which provides a more stable and clinically interpretable binary outcome than the stricter delta-SLEDAI worsening target. To improve robustness, the final pipeline combines individual gene features with biologically informed module scores, including an interferon signature score, and uses class-weighted models to handle class imbalance without discarding training observations.

## Limitations

\[WRITE HERE\]

Important limitations include repeated patient visits, class imbalance, possible batch effects between GEO datasets, cohort differences, and the use of a binary SLEDAI threshold. External validation partly addresses generalisability by testing whether a model trained on one cohort transfers to another independent cohort.

## Conclusion

\[WRITE HERE\]

Overall, this model evaluates whether current gene expression, immune-module activity, and recent clinical trajectory can predict next-visit SLE disease activity. Strong internal performance would suggest a useful predictive signal, while weaker external performance would indicate the need for additional batch correction, pathway-level modelling, and larger validation cohorts.
